Interisland movements and oceanic swimming of woodland caribou in Newfoundland

### 2020-05-14 paper/manuscript.Rmd
- moved 'local observations' higher in para before discussing GPS analyses - expanded question (1) swimming vs. ice-walking at end

The Fogo Island archipelago, off the coast of Newfoundland, Canada, is home to
approximately 300 woodland caribou. During routine fieldwork on May 30, 2017, we
observed an unmarked adult male caribou swim between two small islands (Figure
1, see Figure 2 for location), a distance of at least 470 m which took ~9
minutes (~52 m per minute). This observation prompted us to investigate whether
swimming behaviour was common among GPS radio-collared caribou in the population
[n = 29, for collaring details see, @Peignier2019]. We identified an
individual's swimming events as two consecutive GPS locations (2-hour relocation
rates)  on different islands. Northeastern Newfoundland typically experiences
pack ice during winter and caribou are known locally to travel between islands
by walking over the ice. We restricted our GPS data to the ice-free period of
the year (April 1 to December 31) and identified `r nevents` swimming events
over three years (Figure 2). In addition to our own visual and remotely sensed
observations, residents of Fogo Island have also reported regularly observing
caribou swimming between islands.

@@

Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration [@Leblond2016],
to avoid predators [@Bergerud1985], and to access islands during calving
[@Bergerud1990]. Even for adept swimmers like caribou, that have hollow hair
shafts that aid in flotation, the energetic expenditure of swimming for
quadrupedal mammals is significantly higher than walking or running, and
drowning is also possible [@Miller1985]. Despite the abundance of coastal and
island caribou herds, only a handful of studies have documented caribou swimming
in the ocean, and most of these lack observation, but rather deduce swimming
based on presence of caribou on a previously uninhabitated island  [e.g.
@Miller1995; @Miller2002; @Jeffery2007; @Ricca2012]. Caribou may swim between
islands in the ocean for similar reasons they swim in freshwater [@Leblond2016],
mainly concerning predator avoidance or to access new habitats and foraging opportunities. We propose
that one such explanation, the forage limitation hypothesis, is the most likely
explanation for oceanic swimming for caribou that live on islands in the Fogo
Island archipelago.

@@

1) How does the energetic cost or risk of swimming differ from walking on sea
ice? The cost of swimming may be lower than the risk of falling through sea ice
and drowning, so what proportion of inter-island movements are swimming events
compared to walking on ice? The presence and extent of sea ice varies annually
in our study area and we are unable to determine the exact timing or duration of
sea ice from past years. However, we qualitatively delineate the timing of
typical sea ice arrival and departure (Figure 2B) and deduce that although the
majority of inter-island movements occur between April 1 and December 31, some
inter-island movements appear to be caribou walking on sea ice between islands.
Future research should focus on the trade-off between these types of movement.

@@

3) Does reproductive status (presence of a calf-at-heel) or demographic class
(age or sex) influence propensity for oceanic swimming? Our initial observation
was of an adult male, but all collared individuals were adult females. Are
younger caribou more likely to swim due to heightened competition? Are males or
females without a calf-at-heel more likely to swim in the summer given the are
not restricted based on presence of a calf-at-heel, as females are.

>>>>

The Fogo Island archipelago, off the coast of Newfoundland, Canada, is home to
approximately 300 woodland caribou. During routine fieldwork on May 30, 2017, we
observed an unmarked adult male caribou swim between two small islands (Figure
1, see Figure 2 for location), a distance of at least 470 m which took ~9
minutes (~52 m per minute). Residents of Fogo Island confirmed this phenomenon and have also reported regularly observing
caribou swimming between islands. This observation prompted us to investigate whether
swimming behaviour was common among GPS radio-collared caribou in the population
[n = 29, for collaring details see, @Peignier2019]. We identified an
individual's swimming events as two consecutive GPS locations (2-hour relocation
rates)  on different islands. Northeastern Newfoundland typically experiences
pack ice during winter and caribou are known locally to travel between islands
by walking over the ice. We restricted our GPS data to the ice-free period of
the year (April 1 to December 31) and identified `r nevents` swimming events
over three years (Figure 2).

@@

Caribou (*Rangifer tarandus*) are exceptional swimmers. Ample evidence exists
that caribou swim in streams, rivers, and lakes during migration [@Leblond2016],
to avoid predators [@Bergerud1985], and to access islands during calving
[@Bergerud1990]. Even for adept swimmers like caribou, which have hollow hair
shafts that aid in flotation, the energetic expenditure of swimming for
quadrupedal mammals is significantly higher than walking or running, and
drowning is also possible [@Miller1985]. Despite the abundance of coastal and
island caribou herds, only a handful of studies have documented caribou swimming
in the ocean, and most of these lack observation, but rather deduce swimming
based on presence of caribou on a previously uninhabitated island  [e.g.
@Miller1995; @Miller2002; @Jeffery2007; @Ricca2012]. Caribou may swim between
islands in the ocean for similar reasons they swim in freshwater [@Leblond2016],
mainly concerning predator avoidance or to access new habitats and foraging opportunities. We propose
that one such explanation, the forage limitation hypothesis, is the most likely
explanation for oceanic swimming for caribou that live on islands in the Fogo
Island archipelago.

@@

1) How do the energetic cost and risk of swimming differ from walking on sea
ice? Walking over solid ice is likely less energetically expensive than swimming, but may pose a greater risk of falling through depending on the thickness and stability of ice. What proportion of inter-island movements are swimming events compared to walking on ice? The presence and extent of sea ice varies annually in our study area and we are unable to determine the exact timing or duration of sea ice from past years. However, we qualitatively delineate the timing of typical sea ice arrival and departure (Figure 2B) and deduce that although the majority of inter-island movements occur between April 1 and December 31, some inter-island movements appear to be caribou walking on sea ice between islands. Future research on the trade-offs of these types of movement should consider both how the costs associated with each differ, but also how the potential benefits - foraging opportunities - may vary seasonally.

@@

3) Does reproductive status (presence of a calf-at-heel) or demographic class (age or sex) influence propensity for oceanic swimming? Our initial observation was of an adult male, but all collared individuals were adult females. Are younger caribou more likely to swim if they are unable to compete with older individuals for resources? Are males or females without a calf-at-heel more likely to swim in the summer given they are not restricted based on presence of a calf-at-heel, as females are?
